---
title: "Weather data on calendars"
description: "MERRA-2 samples on calendars: heatmap layers, energypal palettes, and recasting across resolutions."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weather data on calendars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
viz <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("energypal", quietly = TRUE)
if (viz) library(ggplot2)
```

## The sample

`merra2_cities` ships three contrasting hourly weather series for 2019 —
Helsinki (high latitude), Lima (equatorial), Sydney (southern
hemisphere) — extracted from NASA's MERRA-2 reanalysis:

```{r}
str(merra2_cities)
```

Instants are stamped at 30 minutes past the hour (the MERRA-2
hourly-mean convention); `datetime_to_timeslice()` floors them into hourly
timeslices, so no preprocessing is needed.

## Datetime data on a calendar

`geom_calendar()` cuts a datetime column to timeslices and aggregates into
tiles. Helsinki wind on the full-resolution `d365_h24` calendar, with
energypal's Global Wind Atlas ramp:

```{r wind-heatmap, fig.width = 8, fig.height = 3.6, eval = viz}
hel <- subset(merra2_cities, city == "Helsinki")
ggplot(hel) +
  geom_calendar(calendar = calendars$d365_h24,
                datetime = "datetime", z = "W50M") +
  energypal::scale_fill_energy_c(palette = "windatlas") +
  scale_x_discrete(breaks = sprintf("d%03d", seq(1, 365, by = 30))) +
  scale_y_discrete(breaks = sprintf("h%02d", seq(0, 23, by = 4))) +
  labs(x = "day of year", y = "hour", fill = "m/s",
       title = "Helsinki wind speed at 50 m, 2019") +
  theme_calendar()
```

The representative-day view compresses the same year into `m12_h24`
(month x hour), and the `by=` argument carries the `city` column through
aggregation so facets work. Times are UTC, so each city's solar noon
sits at its longitude's offset (Lima ~16h, Sydney ~01h — a calendar
with `utc_offset_minutes` shifts this to local time):

```{r solar-heatmap, fig.width = 8, fig.height = 3.2, eval = viz}
ggplot(merra2_cities) +
  geom_calendar(calendar = calendars$m12_h24,
                datetime = "datetime", z = "SWGDN", by = "city") +
  facet_wrap(~city) +
  energypal::scale_fill_energy_c(palette = "solaratlas_ghi") +
  labs(x = "month", y = "hour", fill = "W/m2",
       title = "Mean solar irradiance by month and hour, 2019") +
  theme_calendar()
```

## Timeslice-keyed data

Once data lives on timeslices, `geom_calendar_tile()` draws it directly, and
`join_calendar()` attaches timeframe columns for manual workflows:

```{r timeslice-tiles, fig.width = 7, fig.height = 3, eval = viz}
cal <- calendars$m12_h24
hel$timeslice <- datetime_to_timeslice(hel$datetime, cal)
hel_timeslices <- aggregate(T10M ~ timeslice, data = hel, FUN = mean)

ggplot(hel_timeslices) +
  geom_calendar_tile(calendar = cal, z = "T10M") +
  scale_fill_gradientn(colours = energypal::energypal("solaratlas_ghi")) +
  labs(x = "month", y = "hour", fill = "degC",
       title = "Helsinki mean temperature (m12_h24)") +
  theme_calendar()
```

(energypal deliberately ships no temperature palette; the solar-atlas
ramp is borrowed here and labeled as such.)

## Recasting across calendars

Timeslice-keyed panels recast between any two calendars, and identifier
columns like `city` are preserved as grouping columns — the same
mechanics that let mixed pipelines chain
`x |> recast_calendar(...) |> geo_recast(...)`. Aggregate all three
cities onto `m12_h24`, then move the panel across resolutions:

```{r recast-panel}
all3 <- merra2_cities
all3$timeslice <- datetime_to_timeslice(all3$datetime, cal)
panel <- aggregate(cbind(T10M, W50M) ~ city + timeslice, data = all3,
                   FUN = mean)

# month x hour -> quarter x hourtype (cross-cutting hierarchies)
q_hp <- recast_calendar(panel, cal, calendars$q4_hp3, year = 2019,
                        rule = "weighted_mean")
head(q_hp, 6)

# ... and to the seasonal view
seasonal <- recast_calendar(panel, cal, calendars$s4, year = 2019,
                            rule = "weighted_mean")
seasonal[seasonal$city == "Helsinki", ]
```

Conservation under `rule = "sum"` holds per identifier group. Treating
mean irradiance as an extensive proxy (energy per timeslice), each city's
annual total survives the trip to `"ANNUAL"`:

```{r conservation}
irr <- aggregate(SWGDN ~ city + timeslice, data = all3, FUN = sum)
annual <- recast_calendar(irr, cal, to = "ANNUAL", year = 2019,
                          rule = "sum")
annual
# identical to summing the timeslice panel directly:
tapply(irr$SWGDN, irr$city, sum)
```

## Comparing resolutions

The same temperature series at monthly resolution for all three cities,
with `energypal_gradient()` spreading one carrier color into related
tones (the levels aren't energy carriers, so explicit colors sidestep
the label matcher):

```{r resolutions, fig.width = 7, fig.height = 3.4, eval = viz}
m_t <- recast_calendar(panel[c("city", "timeslice", "T10M")], cal,
                       to = "MONTH", year = 2019, rule = "weighted_mean")

tones <- energypal::energypal_gradient(
  energypal::energypal("carriers")[["Solar"]], n = 3)

ggplot(m_t, aes(timeslice, T10M, color = city, group = city)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = unname(tones)) +
  labs(x = NULL, y = "degC",
       title = "Monthly mean temperature, 2019") +
  theme_calendar()
```
