---
title: "Calendar catalog"
description: "The built-in calendar designs: what ships, how to build them, and how to pick one."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calendar catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
library(dplyr, warn.conflicts = FALSE)
gg <- requireNamespace("ggplot2", quietly = TRUE)
```

## The catalog

`timescales` ships a curated catalog of named calendar designs, ported
from the predecessor `timeslices` package. Every id can be built fresh
with `calendar(id)`, and the same objects come pre-built in the
`calendars` dataset:

```{r}
cal <- calendar("m12_h24")     # build on demand
cal2 <- calendars$m12_h24      # pre-built package data
identical(calendar_leaftable(cal), calendar_leaftable(cal2))
```

**Coverage** says how the design relates to a real year: `complete`
(every instant has a timeslice), `truncated` (a stylised year that drops
instants — `d365` has no Feb 29), or `representative` (timeslices stand for
recurring types rather than a partition of the timeline — `q4_h24` is a
representative day per quarter). **Regularity** flags designs whose
timeslices differ in length within a level (`m12_md365`: February is short).

The full table is one `calendar_catalog()` call; below it comes in
families, each with one member drawn as an icicle (`autoplot()`; widths
are the timeslice shares, the `ANNUAL` root on top).

```{r catalog-helpers, include = FALSE}
cat_df <- calendar_catalog() |>
  select(id, timeframes, n_timeslices, coverage_class, regularity)
fam <- function(pattern) {
  cat_df |>
    filter(grepl(pattern, id)) |>
    knitr::kable(row.names = FALSE)
}
```

### Days of the year

Plain day counters, optionally with hours: the full-resolution
workhorses (`d365_h24` is the classic 8,760) and the stylised 360/364-day
years some models use.

```{r}
fam("^d3[0-9]{2}(_h24)?$")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$d365_h24)
```

### Months

Twelve calendar months (`m01..m12`, or `JAN..DEC` as `m12a`), optionally
with a representative day of 24 hours.

```{r}
fam("^m12(a|_h24|a_h24)?$")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$m12_h24)
```

### Month x day-of-month

A real month/day structure — non-Cartesian, February is short — for data
keyed by month and day-of-month.

```{r}
fam("^m12_md")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$m12_md365)
```

### Quarters and seasons

Four calendar quarters, or the meteorological seasons (`WIN` = Dec-Feb —
a season straddles the year boundary, which is why it is its own axis).

```{r}
fam("^(q4|s4)(_h24)?$")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$s4_h24)
```

### Weeks

52 or 53 weeks, with a representative day (`_h24`) or the full 168-hour
week (`_h168`).

```{r}
fam("^w5")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$w52_h24)
```

### Weekdays and day types

The seven weekdays, or the compact workday/weekend pair — the smallest
designs that still resolve a weekly demand cycle.

```{r}
fam("^(wd7|wk2)")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$wk2_h24)
```

### Hour types

`DAY`/`NIGHT`/`PEAK` hour classes on their own or under a coarser axis —
the "typical periods" family.

```{r}
fam("hp3$")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$s4_hp3)
```

### Fiscal years (April-start)

The same designs anchored to April 1 — Indian fiscal reporting; note the
member order starting at `m04`/`Q2` (details in the
[Fiscal years](#fiscal-years-the-fy04_-family) section below).

```{r}
fam("^fy04")
```

```{r, fig.width = 8, fig.height = 1.9, eval = gg}
ggplot2::autoplot(calendars$fy04_m12)
```

## Shares are duration-proportional

Unlike the timeslices originals — which gave every timeslice a uniform
share — catalog calendars weight timeslices by real duration:

```{r}
m12 <- calendars$m12
data.frame(timeslice = calendar_leaftable(m12)$timeslice, share = round(calendar_leaftable(m12)$share, 4))[1:3, ]
```

January is `31/365` of the year, not `1/12`. This is what makes
`recast_calendar(rule = "sum")` conserve and `weighted_mean` weight correctly.

## The workhorses

The designs most models start from:

```{r}
calendars$d365_h24    # 8,760 hourly timeslices, the full-resolution classic
calendars$m12_h24     # 288 timeslices: representative day per month
calendars$q4_h24      # 96 timeslices: representative day per quarter
```

The `m12_md*` family carries a real month/day structure (non-Cartesian —
February is short), useful when data arrives keyed by month and
day-of-month:

```{r}
md <- calendars$m12_md365
nrow(calendar_leaftable(md))
head(calendar_leaftable(md)[calendar_leaftable(md)$MONTH == "m02", ], 2)
tail(calendar_leaftable(md)[calendar_leaftable(md)$MONTH == "m02", ], 1)   # Feb ends at d28
datetime_to_timeslice(as.Date(c("2021-03-15", "2020-02-29")), md)  # Feb 29 -> NA
```

## Type axes: SEASON, DAYTYPE, HOURTYPE

Three derived timeframes back the compact "typical periods" designs —
and unlike in timeslices, they are fully datetime-convertible:

- `SEASON` — meteorological: `WIN` = Dec–Feb, `SPR`, `SUM`, `FAL`
- `DAYTYPE` — `WORKDAY` (Mon–Fri) / `WEEKEND`
- `HOURTYPE` — `NIGHT` (h22–h05), `PEAK` (h17–h20), `DAY` (the rest)

```{r}
dtm <- as.POSIXct(c("2021-01-15 03:00", "2021-01-16 18:00",
                    "2021-07-14 12:00"), tz = "UTC")
datetime_to_timeslice(dtm, calendars$s4_hp3)
datetime_to_timeslice(dtm, calendars$wk2_h24)
```

The mappings are defaults, not dogma — register your own token (e.g. a
different peak window) with `register_calendar_token()` and build a custom
calendar from it.

## Fiscal years: the `fy04_*` family

Many reporting systems run April..March — Indian national statistics
most prominently. The `fy04_*` entries (`fy04_m12`, `fy04_m12_h24`,
`fy04_q4`, `fy04_q4_h24`, `fy04_d365`, `fy04_d365_h24`) anchor the model
year to April 1: model year `y` spans `[y-04-01, y+1-04-01)`, and the
anchored `YEAR` is the *starting* Gregorian year (Indian "FY 2021-22" is
model year 2021).

```{r}
fy <- calendar("fy04_m12")
fy@members$MONTH                       # April first
g <- expand_calendar(fy, 2021, by = "day")
range(as.Date(g$datetime))             # 2021-04-01 .. 2022-03-31
datetime_to_timeslice(as.Date("2022-01-15"), fy)   # January stays m01
```

Two rules keep fiscal calendars unambiguous:

- **Labels stay Gregorian.** April is `m04` and Apr–Jun is `Q2`, always;
  only the member *order* starts at the anchor (`m04..m03`,
  `Q2,Q3,Q4,Q1`). Fiscal-renumbered labels would silently collide with
  the Gregorian label matcher.
- **The catalog is UTC.** Data already in Indian local time maps as-is;
  to map true-UTC instants at Indian midnight boundaries, pass the IST
  offset: `calendar("fy04_m12", utc_offset_minutes = 330L)`. (Constant
  offsets only for now — Olson time zones / DST are a planned later
  phase.)

Leap years take care of themselves: `fy04_d365` drops the Feb 29 that
falls in the *following* Gregorian year (FY2023 contains 2024-02-29) and
keeps March aligned to `d365`. Any other anchor works by argument on any
design: `calendar("m12", year_start = list(month = 7L, day = 1L))`.

## Visualizing calendars

(For the full plot-type tour on real data -- heatmap layers, wall
calendars, profiles, ribbons, duration curves -- see
[Visualization](visualization.html).)

`autoplot()` (or `plot()`) draws a calendar's structure as an icicle: one
band per timeframe with the `ANNUAL` root on top, rectangle widths equal
to timeslice shares, x spanning the year on `[0, 1]`. The gradient
restarts inside each parent (`color_pattern = "within"`), so nested
structure is visible at a glance — the catalog tour above shows one per
family. Dense calendars stay fast: rows beyond `max_segments` are binned
before drawing:

```{r, fig.width = 8, fig.height = 2.8, eval = gg}
library(ggplot2)
autoplot(calendars$d365_h24, max_segments = 1000)
```

The same figure carries data: pass `data =`/`z =` and every band fills
with the value recast to that band's resolution — here a year of
Reykjavik wind, from the annual mean down to the month × hour grid:

```{r icicle-data, fig.width = 8, fig.height = 2.6, eval = gg}
wind <- merra2_cities |>
  filter(city == "Reykjavik") |>
  mutate(timeslice = datetime_to_timeslice(datetime, calendars$m12_h24)) |>
  summarise(W50M = mean(W50M), .by = timeslice)

autoplot(calendars$m12_h24, data = wind, z = "W50M",
         rule = "weighted_mean", year = 2019) +
  labs(fill = "m/s")
```

The geometry itself is available without ggplot2 via `calendar_layout()`
— a plain `data.frame` of rectangles, usable from any plotting system:

```{r}
head(calendar_layout(calendars$q4_h24), 7)
```

`calendar_plot()` is the data-on-calendar heatmap: hand it a `data.frame`
keyed by timeslice (or nothing, to see the share structure). The layout
follows the hierarchy — finest timeframe on y, next on x, coarser levels
as facets:

```{r, fig.width = 7, fig.height = 4, eval = requireNamespace("ggplot2", quietly = TRUE)}
x <- data.frame(timeslice = calendar_leaftable(calendars$m12_h24)$timeslice)
x$load <- 80 + 40 * sin(seq(0, 6 * pi, length.out = nrow(x)))
calendar_plot(calendars$m12_h24, x, palette = "C")
```

## Wall calendars

`calendar_wall_plot()` draws the familiar wall form: one facet per
month, day cells in a week grid. The weekday arrangement is
year-specific (which weekday a date falls on comes from the base grid),
so pass `year=`; without it the layout falls back to a year-free
7-wide sequence with a message.

```{r, fig.width = 8, fig.height = 6}
calendar_wall_plot(calendar("m12_md365"), year = 2021)
```

Daily (or finer -- aggregated by `fun`) data fills the cells, and the
fiscal calendars facet April-first out of the box:

```{r, fig.width = 8, fig.height = 6}
cal <- calendar("fy04_d365")
x <- data.frame(timeslice = calendar_leaftable(cal)$timeslice,
                v = cumsum(rnorm(365)))
calendar_wall_plot(cal, x, z = "v", year = 2021)
```

The pieces are reusable on their own: `calendar_weekdays(cal, year,
week_start = "MON")` tabulates the day layer's dates, weekdays,
week-of-month rows and an anchored week-of-year (fiscal weeks from
April 1 for `fy04_*`; `week_start` accepts any of `MON..SUN`), and
`calendar_wall_layout()` returns the plain tile frame the figure is
built from.

## Recasting across the catalog

Any two catalog calendars convert into each other; totals are conserved
under `rule = "sum"`:

```{r}
x <- data.frame(timeslice = calendar_leaftable(calendars$m12)$timeslice,
                energy = c(310, 280, 300, 250, 220, 230,
                           260, 270, 240, 250, 280, 320))
recast_calendar(x, calendars$m12, calendars$s4, year = 2021, rule = "sum",
       by = "day")
sum(x$energy)
```

Or aggregate within one calendar via the `ANNUAL` root:

```{r}
recast_calendar(x, calendars$m12, to = "ANNUAL", year = 2021, rule = "sum",
       by = "day")
```
