---
title: "Calendar catalog"
description: "The built-in calendar designs: what ships, how to build them, and how to pick one."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calendar catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
```

## The catalog

`timescales` ships a curated catalog of named calendar designs, ported
from the predecessor `timeslices` package. Every id can be built fresh
with `calendar(id)`, and the same objects come pre-built in the
`calendars` dataset:

```{r}
knitr::kable(calendar_catalog())
```

Two equivalent ways to get one:

```{r}
cal <- calendar("m12_h24")     # build on demand
cal2 <- calendars$m12_h24      # pre-built package data
identical(cal@leaves, cal2@leaves)
```

**Coverage** says how the design relates to a real year: `complete`
(every instant has a slice), `truncated` (a stylised year that drops
instants — `d365` has no Feb 29), or `representative` (slices stand for
recurring types rather than a partition of the timeline — `q4_h24` is a
representative day per quarter). **Regularity** flags designs whose
slices differ in length within a level (`m12_md365`: February is short).

## Shares are duration-proportional

Unlike the timeslices originals — which gave every slice a uniform
share — catalog calendars weight slices by real duration:

```{r}
m12 <- calendars$m12
data.frame(slice = m12@leaves$slice, share = round(m12@leaves$share, 4))[1:3, ]
```

January is `31/365` of the year, not `1/12`. This is what makes
`recast(rule = "sum")` conserve and `weighted_mean` weight correctly.

## The workhorses

The designs most models start from:

```{r}
calendars$d365_h24    # 8,760 hourly slices, the full-resolution classic
calendars$m12_h24     # 288 slices: representative day per month
calendars$q4_h24      # 96 slices: representative day per quarter
```

The `m12_md*` family carries a real month/day structure (non-Cartesian —
February is short), useful when data arrives keyed by month and
day-of-month:

```{r}
md <- calendars$m12_md365
nrow(md@leaves)
head(md@leaves[md@leaves$MONTH == "m02", ], 2)
tail(md@leaves[md@leaves$MONTH == "m02", ], 1)   # Feb ends at d28
instant_to_slice(as.Date(c("2021-03-15", "2020-02-29")), md)  # Feb 29 -> NA
```

## Type axes: SEASON, DAYTYPE, HOURTYPE

Three derived timeframes back the compact "typical periods" designs —
and unlike in timeslices, they are fully datetime-convertible:

- `SEASON` — meteorological: `WIN` = Dec–Feb, `SPR`, `SUM`, `FAL`
- `DAYTYPE` — `WORKDAY` (Mon–Fri) / `WEEKEND`
- `HOURTYPE` — `NIGHT` (h22–h05), `PEAK` (h17–h20), `DAY` (the rest)

```{r}
dtm <- as.POSIXct(c("2021-01-15 03:00", "2021-01-16 18:00",
                    "2021-07-14 12:00"), tz = "UTC")
instant_to_slice(dtm, calendars$s4_hp3)
instant_to_slice(dtm, calendars$wk2_h24)
```

The mappings are defaults, not dogma — register your own token (e.g. a
different peak window) with `register_token()` and build a custom
calendar from it.

## Visualizing calendars

`autoplot()` (or `plot()`) draws a calendar's structure as an icicle: one
band per timeframe with the `ANNUAL` root on top, rectangle widths equal
to slice shares, x spanning the year on `[0, 1]`. The gradient restarts
inside each parent (`color_pattern = "within"`), so nested structure is
visible at a glance:

```{r, fig.width = 8, fig.height = 2.8, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
autoplot(calendars$m12_h24)
autoplot(calendars$s4_hp3)
autoplot(calendars$m12_md365)
```

Dense calendars stay fast — rows beyond `max_segments` are binned before
drawing:

```{r, fig.width = 8, fig.height = 2.8, eval = requireNamespace("ggplot2", quietly = TRUE)}
autoplot(calendars$d365_h24)
```

The geometry itself is available without ggplot2 via `calendar_layout()`
— a plain `data.frame` of rectangles, usable from any plotting system:

```{r}
head(calendar_layout(calendars$q4_h24), 7)
```

`calendar_plot()` is the data-on-calendar heatmap: hand it a `data.frame`
keyed by slice (or nothing, to see the share structure). The layout
follows the hierarchy — finest timeframe on y, next on x, coarser levels
as facets:

```{r, fig.width = 7, fig.height = 4, eval = requireNamespace("ggplot2", quietly = TRUE)}
x <- data.frame(slice = calendars$m12_h24@leaves$slice)
x$load <- 80 + 40 * sin(seq(0, 6 * pi, length.out = nrow(x)))
calendar_plot(calendars$m12_h24, x, palette = "C")
```

## Recasting across the catalog

Any two catalog calendars convert into each other; totals are conserved
under `rule = "sum"`:

```{r}
x <- data.frame(slice = calendars$m12@leaves$slice,
                energy = c(310, 280, 300, 250, 220, 230,
                           260, 270, 240, 250, 280, 320))
recast(x, calendars$m12, calendars$s4, year = 2021, rule = "sum",
       by = "day")
sum(x$energy)
```

Or aggregate within one calendar via the `ANNUAL` root:

```{r}
recast(x, calendars$m12, to = "ANNUAL", year = 2021, rule = "sum",
       by = "day")
```
