---
title: "Concepts"
description: "The mental model: timeframes, calendars, leaves, tokens, recasting."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concepts}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
```

## Three layers of abstraction

`timescales` is built around three concentric ideas. Working out from
the centre:

1. **Timeframes** — the atomic vocabulary (`YEAR`, `MONTH`, `HOUR`, ...).
2. **Tokens** — named recipes for the labels at one timeframe
   (e.g. `m12` = `m01..m12`, `m12a` = `JAN..DEC`).
3. **Calendars** — the actual partition of a year into labelled
   leaves, formed by composing tokens.

Most users only ever touch layer 3.

## 1. Timeframes

A *timeframe* is a unit of time at which a calendar can carry labels.
The package ships with a fixed core set:

```{r}
CORE_TIMEFRAMES
```

The function `as_timeframe()` extracts the value of a chosen timeframe
from any datetime:

```{r}
t <- as.POSIXct("2025-04-15 13:45", tz = "UTC")
as_timeframe(t, "MONTH")
as_timeframe(t, "MONTH", format = "token")
as_timeframe(t, "HOUR")
as_timeframe(t, "WDAY", format = "token")
```

`format = "token"` returns the same label style that the built-in
tokens use, which is what the calendar machinery needs internally.

## 2. Tokens

A *token* is a named, reusable recipe for the labels of one timeframe,
plus each label's within-year share. Tokens are how vocabularies are
shared across calendars.

```{r}
list_tokens()
get_token("m12")$expand() |> head(3)
get_token("h24")$expand() |> head(3)
```

A few built-ins:

| Token   | Timeframe | Labels              | Notes                           |
|---------|-----------|---------------------|---------------------------------|
| `d365`  | YDAY      | `d001..d365`        | day-of-year, 365-day calendar   |
| `m12`   | MONTH     | `m01..m12`          | numeric month                   |
| `m12a`  | MONTH     | `JAN..DEC`          | abbreviated month               |
| `q4`    | QUARTER   | `Q1..Q4`            | day-weighted shares             |
| `wd7`   | WDAY      | `MON..SUN`          | ISO weekday order               |
| `h24`   | HOUR      | `h00..h23`          | hour-of-day                     |
| `h168`  | HOUR      | `h000..h167`        | hour-of-week                    |

Custom tokens are added with `register_token()`.

## 3. Calendars

A **Calendar** is the actual product of one or more tokens. The leaves
of the resulting tree are the slices the rest of your model talks about.

Three constructors, in increasing flexibility:

```{r, eval = FALSE}
calendar("m12_h24")                    # layer 1: by name
calendar_build("m12", "h24")           # layer 2: by token list
calendar_from_leaves(leaves, ...)      # layer 3: by raw leaf table
```

The first two compose tokens with a Cartesian product; share is the
product of per-token shares scaled to `year_fraction`. The third is
the escape hatch for irregular calendars (e.g. representative weeks
that intentionally do not cover the full year).

### What the leaves table looks like

```{r}
cal <- calendar("q4_h24")
head(cal@leaves, 3)
cat(nrow(cal@leaves), "leaves, share sums to", sum(cal@leaves$share))
```

## 4. Recasting between calendars

Conversion is the central verb. The same `recast()` function handles
upsampling, downsampling, and irregular-to-irregular mappings, given
that both calendars cover the same year fraction.

```{r}
cal_m <- calendar("m12")     # source: monthly
cal_q <- calendar("q4")      # target: quarterly

monthly <- data.frame(
  slice = cal_m@leaves$slice,
  load  = c(120, 118, 105,  92,  85,  88,  95, 100,  98,  90, 105, 122)
)

recast(monthly, from = cal_m, to = cal_q, year = 2025,
       rule = "weighted_mean", by = "day")
```

### Aggregation rules

| Rule              | Meaning                                                        |
|-------------------|----------------------------------------------------------------|
| `weighted_mean`   | Average weighted by leaf share (the default).                  |
| `mean`            | Plain mean over the shared expansion grid.                     |
| `sum`             | Sum over the shared expansion grid.                            |

On a uniform grid `weighted_mean` and `mean` coincide because longer
source slices contribute proportionally more rows.

## Design boundaries

- **Year-bounded.** A calendar describes one year (or a fraction of one).
  Multi-year horizons are out of scope for v0.1; build them on top of
  this layer.
- **Label-stable.** A calendar's slice IDs and ordering are fixed at
  construction time. Conversions never mutate them.
- **No side effects.** Every constructor returns a value; nothing is
  registered or globally configured.
