---
title: "Concepts"
description: "The mental model: timeframes, calendars, leaves, tokens, recasting."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
```

## Three layers of abstraction

`timescales` is built around three concentric ideas. Working out from
the centre:

1. **Timeframes** — the atomic vocabulary (`YEAR`, `MONTH`, `HOUR`, ...).
2. **Tokens** — named recipes for the labels at one timeframe
   (e.g. `m12` = `m01..m12`, `m12a` = `JAN..DEC`).
3. **Calendars** — the actual partition of a year into labelled
   leaves, formed by composing tokens.

Most users only ever touch layer 3.

## 1. Timeframes

A *timeframe* is a unit of time at which a calendar can carry labels.
The package ships with a fixed core set:

```{r}
CORE_TIMEFRAMES
```

The function `as_timeframe()` extracts the value of a chosen timeframe
from any datetime:

```{r}
t <- as.POSIXct("2025-04-15 13:45", tz = "UTC")
as_timeframe(t, "MONTH")
as_timeframe(t, "MONTH", format = "token")
as_timeframe(t, "HOUR")
as_timeframe(t, "WDAY", format = "token")
```

`format = "token"` returns the same label style that the built-in
tokens use, which is what the calendar machinery needs internally.

## 2. Tokens

A *token* is a named, reusable recipe for the labels of one timeframe,
plus each label's within-year share. Tokens are how vocabularies are
shared across calendars.

```{r}
list_tokens()
get_token("m12")$expand() |> head(3)
get_token("h24")$expand() |> head(3)
```

A few built-ins:

| Token   | Timeframe | Labels              | Notes                           |
|---------|-----------|---------------------|---------------------------------|
| `d365`  | YDAY      | `d001..d365`        | day-of-year; aligns `drop_feb29` |
| `d360`  | YDAY      | `d001..d360`        | stylised year; aligns `drop_last` |
| `m12`   | MONTH     | `m01..m12`          | numeric month                   |
| `m12a`  | MONTH     | `JAN..DEC`          | abbreviated month               |
| `q4`    | QUARTER   | `Q1..Q4`            | day-weighted shares             |
| `wd7`   | WDAY      | `MON..SUN`          | ISO weekday order               |
| `h24`   | HOUR      | `h00..h23`          | hour-of-day                     |
| `h168`  | WHOUR     | `h000..h167`        | hour-of-week (Mon 00:00 = `h000`) |

Custom tokens are added with `register_token()`, optionally declaring an
*alignment* rule (see below).

### Alignment: mapping real years onto stylised ones

A `d365` calendar has no label for Feb 29; a `d360` calendar has none for
the last days of December. *Alignment rules* (`ALIGNMENT_RULES`) declare
what happens to such instants: `drop_feb29` (Feb 29 is `NA`, later ydays
shift down so Dec 31 is still `d365`), `drop_last`, `repeat_last` (clamp
to the last label — how week 53 folds into `w52`), and `exact` (error).
Built-in tokens carry sensible defaults; calendars record them in
`meta$alignment`, and `instant_to_slice()` accepts an override.

```{r}
d365 <- calendar_build("d365")
instant_to_slice(as.Date(c("2020-02-29", "2020-12-31")), d365)
```

## 3. Calendars

A **Calendar** is the actual product of one or more tokens. The leaves
of the resulting tree are the slices the rest of your model talks about.

Three constructors, in increasing flexibility:

```{r, eval = FALSE}
calendar("m12_h24")                    # layer 1: by name
calendar_build("m12", "h24")           # layer 2: by token list
calendar_from_leaves(leaves, ...)      # layer 3: by raw leaf table
```

The first two compose tokens with a Cartesian product; share is the
product of per-token shares scaled to `year_fraction`. The third is
the escape hatch for irregular calendars (e.g. representative weeks
that intentionally do not cover the full year).

### What the leaves table looks like

```{r}
cal <- calendar("q4_h24")
head(cal@leaves, 3)
cat(nrow(cal@leaves), "leaves, share sums to", sum(cal@leaves$share))
```

## 4. Recasting between calendars

Conversion is the central verb. The same `calendar_recast()` function handles
upsampling, downsampling, and irregular-to-irregular mappings, given
that both calendars cover the same year fraction.

```{r}
cal_m <- calendar("m12")     # source: monthly
cal_q <- calendar("q4")      # target: quarterly

monthly <- data.frame(
  slice = cal_m@leaves$slice,
  load  = c(120, 118, 105,  92,  85,  88,  95, 100,  98,  90, 105, 122)
)

calendar_recast(monthly, from = cal_m, to = cal_q, year = 2025,
       rule = "weighted_mean", by = "day")
```

### Aggregation rules

Conversion routes through the **base grid** of real instants: source
values are projected *down* to instants, then aggregated *up* to target
slices, so aggregation and disaggregation are one operation
(`RECAST_RULES`):

| Rule              | Down (slice → instants)      | Up (instants → slice)                    |
|-------------------|------------------------------|------------------------------------------|
| `weighted_mean`   | copy                         | mean weighted by declared `share` (default) |
| `sum`             | split across grid instants   | sum — **totals are conserved**           |
| `mean`            | copy                         | plain (time-weighted) mean               |
| `copy`            | copy                         | the common value; error if not constant  |
| `sd`              | copy                         | spread of the fine signal                |

`weighted_mean` and `mean` differ exactly when a calendar's declared
shares differ from its real-time coverage. Per-parameter defaults can be
registered with `register_rule()`, and a pairwise calendar-to-calendar
override with `register_conversion()`. Instants not covered by one of
the calendars are governed by `na_action = "drop"` (warns), `"error"`,
or `"keep"` (an explicit `NA` row that conserves totals).

### Within-calendar aggregation and the ANNUAL root

Every calendar has an implicit whole-year root named `ANNUAL`.
`calendar_at_level()` truncates a calendar at one of its own timeframes,
and `calendar_recast()` accepts a timeframe name for `to=`:

```{r}
cal <- calendar("q4_h24")
x <- data.frame(slice = cal@leaves$slice, energy = 1)
calendar_recast(x, cal, to = "ANNUAL", year = 2025, rule = "sum")
```

## Design boundaries

- **Year-bounded calendars, multi-year base.** A calendar describes one
  model year (or a fraction of one); the base instant grid
  (`base_calendar()`) spans real multi-year time, which is how leap
  years stay representable. Multi-year horizons remain a future layer.
- **Label-stable.** A calendar's slice IDs and ordering are fixed at
  construction time. Conversions never mutate them.
- **Explicit registries only.** Constructors return values; the token,
  rule, and conversion registries change behaviour only when you
  register into them.
