---
title: "Data manipulation with calendars"
description: "The tidy workflow end to end: map datetimes to timeslices, attach calendars to tables, recast between resolutions, inspect the crosswalk, and run the same pipelines over data.table and arrow backends."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data manipulation with calendars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
library(dplyr, warn.conflicts = FALSE)
has_dt <- requireNamespace("data.table", quietly = TRUE) &&
  requireNamespace("dtplyr", quietly = TRUE)
```

## The toolkit at a glance

| verb | direction | what it does |
|---|---|---|
| `datetime_to_timeslice()` | datetimes → labels | cut a datetime vector to a calendar's timeslice IDs |
| `expand_calendar()` | calendar → grid | enumerate the base datetime grid of a model year |
| `join_calendar()` | calendar → columns | *attach* labels, timeframe columns, share/weight to a table (no aggregation) |
| `recast_calendar()` | calendar A → calendar B | *convert* values between two resolutions, one rule per column |
| `recast_to_timebase()` / `recast_from_timebase()` | the route halves | project down to the base grid / aggregate up from it |
| `calendar_map()` | A → B crosswalk | the conversion, materialised as a small table |
| `register_calendar_rule()` / `register_calendar_map()` / `register_calendar_conversion()` | registries | per-column rules, exact crosswalks, functional overrides |

Everything below runs on the shipped `merra2_cities` sample (three
cities × 8,760 hours of 2019) in tidyverse style — data flows through
`|>`, and every verb accepts a `data.frame`, tibble, `data.table`,
dtplyr, or arrow input (see [Backends](#backends)).

## From datetimes to timeslices

`datetime_to_timeslice()` is the entry point: it applies the calendar's
alignment rules and returns the timeslice ID each instant falls in
(instants the calendar does not cover come back `NA`):

```{r}
cal <- calendars$m12_h24

merra2_cities |>
  mutate(timeslice = datetime_to_timeslice(datetime, cal)) |>
  select(city, datetime, timeslice, T10M) |>
  head(3)
```

The inverse view is `expand_calendar()`: one row per base-grid point of
a model year, with the timeslice that point belongs to — the
year-aware, leap-aware grid every conversion routes through:

```{r}
expand_calendar(calendars$m12, year = 2020, by = "day") |> head(3)
```

## Attaching a calendar to a table

`join_calendar()` *decorates* rather than converts: it adds a label
column **named after the calendar**, plus optional timeframe columns and
share/weight, all `"<name>."`-prefixed. Keys are auto-detected — a
column named like the calendar, else `timeslice`, else a `datetime`
column (labels then computed on the base grid):

```{r}
panel <- merra2_cities |>
  mutate(timeslice = datetime_to_timeslice(datetime, cal)) |>
  summarise(across(c(T10M, SWGDN), mean), .by = c(city, timeslice))

panel |>
  join_calendar(cal, timeframes = TRUE, meta = TRUE) |>
  head(3)
```

Because every calendar attaches under its own name, several can coexist
on one dataset — and a table carrying two label columns is itself an
empirical crosswalk between those calendars:

```{r}
two <- merra2_cities |>
  filter(city == "Helsinki") |>
  join_calendar(calendars$m12_h24) |>
  join_calendar(calendars$q4_hp3)
two |> select(datetime, m12_h24, q4_hp3) |> head(3)
```

Existing columns are never overwritten — a clashing attach errors
instead.

## Recasting between calendars

`recast_calendar()` converts values from one calendar to another —
aggregation and disaggregation are the same operation, routed
`A -> base -> B` through the shared grid. Three things to know:

**1. One rule per value column, and the rule is mandatory.** Pass
`rule=` for all columns, or register per-column rules once; a column
with neither errors — a silently guessed rule is a silent unit error.

```{r}
register_calendar_rule("T10M",  "weighted_mean")   # intensive: temperature
register_calendar_rule("SWGDN", "sum")             # extensive proxy: energy

q <- panel |>
  recast_calendar(cal, calendars$q4, year = 2019)
head(q, 4)
```

| rule | meaning |
|---|---|
| `sum` | conserve totals (split down, sum up) |
| `weighted_mean` | share-weighted mean (declared `share`s) |
| `mean` | plain time-weighted mean over grid points |
| `copy` | constant within the target, error otherwise |
| `sd` | dispersion over grid points (aggregation only) |

**2. Identifier columns ride along.** Columns that are neither the key
nor values (here `city`) are grouping columns, so panel data converts
in one call — and totals conserve *per group*:

```{r}
annual <- panel |>
  select(city, timeslice, SWGDN) |>
  recast_calendar(cal, to = "ANNUAL", year = 2019)
annual

# the same totals, straight from the panel:
panel |> summarise(SWGDN = sum(SWGDN), .by = city)
```

(`to =` also accepts a timeframe name of the source calendar — the
within-calendar aggregation shortcut, `"ANNUAL"` being the root.)

**3. Coverage is explicit.** Grid points the target does not cover are
dropped with a warning by default; `na_action = "keep"` retains them as
an explicit `NA`-timeslice row so totals conserve, and
`na_action = "error"` refuses. Materialised results always return the
FULL target vocabulary (with `NA` where nothing landed), so downstream
joins see a stable schema.

```{r, include = FALSE}
clear_calendar_rules()
```

## The route halves

`recast_calendar()` is the fused route; its halves are public.
`recast_to_timebase()` projects timeslice data DOWN to the base grid
(extensive columns split so totals conserve; a `weight` column carries
the source shares), and `recast_from_timebase()` aggregates
datetime-keyed rows UP into any calendar — their composition is exactly
`recast_calendar()`:

```{r}
m <- data.frame(timeslice = sprintf("m%02d", 1:12), energy = 1:12)

g <- m |>
  recast_to_timebase(calendars$m12, year = 2019, rule = "sum", by = "day")
head(g, 2)
sum(g$energy)                      # 78 -- conserved on the grid

g |>
  recast_from_timebase(calendars$q4, rule = "sum", by = "day")
```

Use the halves when you need to *do something on the grid* between the
two ends — join weather covariates by datetime, filter a season out,
resample — with conservation guarantees intact.

## The crosswalk, inspectable and overridable

Every recast is a join against a small crosswalk table — one row per
overlapping timeslice pair with grid counts and share weights.
`calendar_map()` exposes it:

```{r}
calendar_map(calendars$m12, calendars$q4, year = 2019) |> head(4)
```

When an exact correspondence is known (hand-audited concordances,
provably nested designs), `register_calendar_map()` short-circuits the
grid derivation for that pair; `register_calendar_conversion()` goes one step
further and replaces the whole conversion with your function. Both are
keyed by calendar names, like the attach.

## Backends

The verbs above are single dplyr pipelines, so the SAME code runs over
an in-memory `data.frame`/tibble, a `data.table` (via dtplyr), or an
arrow Dataset/query. Eager inputs come back in their own class; lazy
inputs return the *uncollected query* unless `collect = TRUE`:

```{r, eval = has_dt}
dt <- data.table::as.data.table(panel)

dt |>
  recast_calendar(cal, calendars$q4, year = 2019, rule = "sum") |>
  class()                          # data.table in, data.table out

lazy <- dtplyr::lazy_dt(dt) |>
  recast_calendar(cal, calendars$q4, year = 2019, rule = "sum")
class(lazy)                        # the query, not the result
head(as.data.frame(dplyr::collect(lazy)), 3)
```

Two contract details worth knowing for lazy sources (dtplyr, arrow):
results carry the *observed* target timeslices only — the
full-vocabulary completion happens on materialisation — and the
calendar side of every join is a small in-memory frame, so an on-disk
arrow dataset is never pulled into memory for the calendar arithmetic.

## Across dimensions

The bare `recast()` generic dispatches on the scale object, so one verb
chains time and space when the sibling package geoscales is loaded:

```{r, eval = FALSE}
x |>
  recast(cal_hourly, cal_monthly, year = 2021, rule = "sum") |>
  recast(gs, to = "r32o11", rule = "sum")     # a geoscales Geoscale
```

Same contract on both sides: explicit rules (never a silent fallback),
identifier columns preserved, backends welcome.

## Where to next?

- [Concepts](concepts.html) — why the route always goes through the
  base grid, and the shared *scales naming lattice.
- [Calendar catalog](calendars.html) — the 43 shipped designs, by
  family.
- [Visualization](visualization.html) — the same pipelines flowing into
  ggplot2.
