---
title: "Getting started with timescales"
description: "A short, end-to-end tour: build a calendar, attach data, convert between calendars."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with timescales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
```

## What problem does this solve?

Energy-system, climate, and operations models all carve the year into
discrete *timeslices*. Different models pick different slicings:
365 days, 12 months, 4 quarters × 24 hours, 168 hours of a representative
week, and so on. The timeslice **labels** are arbitrary, the **shares of a
year** are model-defined, and converting data between slicings is
error-prone.

`timescales` represents any such slicing as a **Calendar**: an ordered
hierarchy of *timeframes* (`YEAR`, `MONTH`, `HOUR`, ...) whose finest
nodes — enumerated in the `@leaftable` — are the timeslices. With one
object you get:

- a stable schema for the timeslice labels and their year-share weights,
- well-defined mappings to and from real datetimes,
- well-defined conversions between any two calendars covering the
  same year fraction,
- ggplot2-ready visualization, from heatmap layers to wall calendars.

## A 5-minute tour

### 1. Build a calendar

The fastest path uses **token-based names** — 43 curated designs ship in
the catalog (`calendar_catalog()`), from `d365_h24` down to typical-period
compressions, including the April-anchored fiscal family
(`calendar("fy04_m12")` — Indian reporting years):

```{r}
cal_my <- calendar("m12_h24")    # 12 months × 24 hours = 288 timeslices
cal_my
```

Equivalent declarative form:

```{r}
cal_my2 <- calendar_build("m12", "h24")
identical(cal_my@timeframes, cal_my2@timeframes)
```

For full control there is the lowest-level constructor
`calendar_from_leaftable()`, where you supply the leaves table directly.

### 2. Inspect the structure

A `Calendar` has four parts:

```{r}
cal_my@timeframes               # the timeframe hierarchy, coarsest first
head(calendar_leaftable(cal_my), 4)          # the leaf table (one row per timeslice)
cal_my@members$MONTH             # ordered label vocabulary per timeframe
cal_my@meta[c("name", "year_fraction")]
```

`@leaftable` is a plain `data.frame`. Each row is one timeslice, with
columns:

- `timeslice` — the unique timeslice ID,
- `share` — fraction of a year,
- `weight` — timeslice weight in hours (default `share * 8760`),
- one column per timeframe — the member label at that level.

### 3. Map real datetimes onto the calendar

```{r}
times <- as.POSIXct(c("2025-01-15 03:00", "2025-07-20 18:00"), tz = "UTC")
datetime_to_timeslice(times, cal_my)
```

### 4. Convert data between calendars

Suppose you have monthly load values and need quarterly averages:

```{r}
cal_m <- calendar("m12")
cal_q <- calendar("q4")

monthly <- data.frame(
  timeslice = sprintf("m%02d", 1:12),
  load  = c(120, 118, 105,  92,  85,  88,  95, 100,  98,  90, 105, 122)
)

recast_calendar(monthly, from = cal_m, to = cal_q, year = 2025,
       rule = "weighted_mean", by = "day")
```

The result is day-weighted: Q1 = (31·v₁ + 28·v₂ + 31·v₃) / 90. The
`rule` is deliberately mandatory — pass one, or register it per column
with `register_calendar_rule()`; a silently guessed rule would be a silent unit
error. `join_calendar()` is the lighter companion: it *attaches* a
calendar's labels and timeframe columns to a dataset instead of
converting it.

### 5. Visualize

Every figure tier works through normal ggplot2 — a heatmap layer here;
wall calendars, profiles, duration curves, and the data-filled
structure figures (icicles and stacks take `data =`/`z =`) in
`vignette("visualization")`:

```{r, fig.width = 7, fig.height = 3, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
x <- data.frame(timeslice = calendar_leaftable(cal_my)$timeslice)
x$load <- 80 + 40 * sin(seq(0, 6 * pi, length.out = nrow(x)))
ggplot(x) +
  geom_calendar_tile(calendar = cal_my, z = "load") +
  scale_fill_viridis_c(option = "H") +
  scale_y_discrete(breaks = calendar_breaks()) +
  labs(x = "month", y = "hour", fill = "load") +
  theme_calendar()
```

## Where to next?

- [Concepts](concepts.html) — the core ideas behind calendars and timeframes.
- [Data structures](data-structures.html) — anatomy of a `Calendar` object
  and its supporting registries.
- [Data manipulation](data-manipulation.html) — the tidy workflow end to
  end: attach, recast, crosswalks, backends.
- [Calendars](calendars.html) — the built-in catalog, fiscal years, and
  wall calendars.
- [Visualization](visualization.html) — the ggplot2 integration contract
  and the full plot-type tour on real weather data.
