---
title: "Visualization with ggplot2"
description: "The timescales viz surface on MERRA-2 weather samples: calendar heatmap layers, wall calendars, profiles, ribbons, duration curves, and the ggplot2 integration contract."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(timescales)
library(dplyr, warn.conflicts = FALSE)
viz <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("energypal", quietly = TRUE)
if (viz) library(ggplot2)
```

## The viz surface at a glance

| kind | functions | use when |
|---|---|---|
| composable layers | `geom_calendar()`, `geom_calendar_tile()`, `theme_calendar()`, `calendar_breaks()` | you are building your own `ggplot()` and want data-on-calendar tiles as one layer among others |
| assembled figures | `calendar_autoplot()` (structure icicle), `calendar_plot()` (heatmap), `calendar_wall_plot()` (wall calendar) | you want a finished figure in one call |
| layout workers | `calendar_layout()`, `calendar_wall_layout()`, `calendar_weekdays()` | you want the plain data frames behind the figures, to draw something the package did not anticipate |

This vignette walks through all three tiers on one running example — and
because every axis is an ordinary vocabulary-ordered factor, everything
composes with standard ggplot2 scales, facets, and themes.

## The sample

`merra2_cities` ships three contrasting hourly weather series for 2019 —
Helsinki (high latitude), Lima (equatorial), Sydney (southern
hemisphere) — extracted from NASA's MERRA-2 reanalysis:

```{r}
str(merra2_cities)
```

Instants are stamped at 30 minutes past the hour (the MERRA-2
hourly-mean convention); `datetime_to_timeslice()` floors them into hourly
timeslices, so no preprocessing is needed.

## The integration contract

The geoms are deliberately *thin*. Each one returns **a single standard
`geom_tile()` layer** whose data is derived from the plot (or layer)
data; nothing is a custom ggproto Geom or Stat. That has three visible
consequences:

* **Calendar inputs are column names, not `aes()` mappings** —
  `datetime =`, `timeslice =`, `z =` name columns of the data. ggplot2
  trains positional scales *before* statistics run, so a Stat cannot
  emit the discrete axes a calendar heatmap needs; a layer factory that
  pre-aggregates can. (The predecessor package solved this with ~340
  lines of custom `scale_x/y_calendar` machinery — deliberately not
  ported.)
* **Axes are vocabulary-ordered factors**, so `scale_x_discrete()`,
  `facet_wrap()`, coordinate flips, and themes all work through the
  normal ggplot2 path. Dense vocabularies thin cleanly with
  `calendar_breaks()`, which always keeps the end values on the axis.
  Fiscal calendars order themselves: an `fy04_*` axis starts at `m04`
  because that is its member order.
* **Facet columns ride through `by =`** — aggregation happens within
  each combination of the listed columns, and those columns survive
  into the layer data, ready for `facet_wrap()`.

## Heatmaps: data on a calendar

`geom_calendar()` cuts a datetime column to timeslices and aggregates into
tiles. Helsinki wind on the full-resolution `d365_h24` calendar, with
energypal's Global Wind Atlas ramp:

```{r wind-heatmap, fig.width = 8, fig.height = 3.6, eval = viz}
hel <- merra2_cities |> filter(city == "Helsinki")

ggplot(hel) +
  geom_calendar(calendar = calendars$d365_h24,
                datetime = "datetime", z = "W50M") +
  energypal::scale_fill_energy_c(palette = "windatlas") +
  scale_x_discrete(breaks = calendar_breaks(10)) +
  scale_y_discrete(breaks = calendar_breaks()) +
  labs(x = "day of year", y = "hour", fill = "m/s",
       title = "Helsinki wind speed at 50 m, 2019") +
  theme_calendar()
```

The representative-day view compresses the same year into `m12_h24`
(month x hour), and the `by=` argument carries the `city` column through
aggregation so facets work. `calendar_breaks()` keeps the twelve-month
and 24-hour axes legible inside narrow facets while pinning the end
values. Times are UTC, so each city's solar noon sits at its longitude's
offset (Lima ~16h, Sydney ~01h — a calendar with `utc_offset_minutes`
shifts this to local time):

```{r solar-heatmap, fig.width = 8, fig.height = 3.2, eval = viz}
ggplot(merra2_cities) +
  geom_calendar(calendar = calendars$m12_h24,
                datetime = "datetime", z = "SWGDN", by = "city") +
  facet_wrap(~city) +
  energypal::scale_fill_energy_c(palette = "solaratlas_ghi") +
  scale_x_discrete(breaks = calendar_breaks(4)) +
  scale_y_discrete(breaks = calendar_breaks()) +
  labs(x = "month", y = "hour", fill = "W/m2",
       title = "Mean solar irradiance by month and hour, 2019") +
  theme_calendar()
```

## Wall calendars

`calendar_wall_plot()` puts the same daily story in the familiar wall
form: one facet per month, day cells in a week grid, weekday initials
across the top. The hourly datetimes are detected automatically and
collapse into day cells with `fun`:

```{r wall, fig.width = 8, fig.height = 6, eval = viz}
calendar_wall_plot(calendar("m12_md365"), hel, z = "T10M",
                   year = 2019) +
  scale_fill_viridis_c(option = "H") +
  labs(fill = "degC",
       title = "Helsinki daily mean temperature, 2019")
```

The weekday arrangement needs `year=` (which weekday a date falls on is
a fact about the year); `week_start` accepts any of `MON..SUN`. See
`vignette("calendars")` for the wall details and `calendar_weekdays()`
for the underlying table.

## Timeslice-keyed data

Once data lives on timeslices, `geom_calendar_tile()` draws it directly, and
`join_calendar()` attaches timeframe columns for manual workflows:

```{r timeslice-tiles, fig.width = 7, fig.height = 3, eval = viz}
cal <- calendars$m12_h24
hel_timeslices <- hel |>
  mutate(timeslice = datetime_to_timeslice(datetime, cal)) |>
  summarise(T10M = mean(T10M), .by = timeslice)

ggplot(hel_timeslices) +
  geom_calendar_tile(calendar = cal, z = "T10M") +
  scale_fill_viridis_c(option = "H") +
  scale_x_discrete(breaks = calendar_breaks()) +
  scale_y_discrete(breaks = calendar_breaks()) +
  labs(x = "month", y = "hour", fill = "degC",
       title = "Helsinki mean temperature (m12_h24)") +
  theme_calendar()
```

## Recasting across calendars

Timeslice-keyed panels recast between any two calendars, and identifier
columns like `city` are preserved as grouping columns — the same
mechanics that let mixed pipelines chain
`x |> recast_calendar(...) |> recast_geoscale(...)`. Aggregate all three
cities onto `m12_h24`, then move the panel across resolutions:

```{r recast-panel}
panel <- merra2_cities |>
  mutate(timeslice = datetime_to_timeslice(datetime, cal)) |>
  summarise(across(c(T10M, W50M), mean), .by = c(city, timeslice))

# month x hour -> quarter x hourtype (cross-cutting hierarchies)
panel |>
  recast_calendar(cal, calendars$q4_hp3, year = 2019,
                  rule = "weighted_mean") |>
  head(6)

# ... and to the seasonal view
panel |>
  recast_calendar(cal, calendars$s4, year = 2019,
                  rule = "weighted_mean") |>
  filter(city == "Helsinki")
```

Conservation under `rule = "sum"` holds per identifier group. Treating
mean irradiance as an extensive proxy (energy per timeslice), each city's
annual total survives the trip to `"ANNUAL"`:

```{r conservation}
irr <- merra2_cities |>
  mutate(timeslice = datetime_to_timeslice(datetime, cal)) |>
  summarise(SWGDN = sum(SWGDN), .by = c(city, timeslice))

irr |>
  recast_calendar(cal, to = "ANNUAL", year = 2019, rule = "sum")

# identical to summing the timeslice panel directly:
irr |> summarise(SWGDN = sum(SWGDN), .by = city)
```

## Profiles: lines over timeslices

Because timeslices come back as ordered factors, line plots need only
`group=`. The same temperature series at monthly resolution for all
three cities:

```{r resolutions, fig.width = 7, fig.height = 3.4, eval = viz}
panel |>
  select(city, timeslice, T10M) |>
  recast_calendar(cal, to = "MONTH", year = 2019,
                  rule = "weighted_mean") |>
  ggplot(aes(timeslice, T10M, color = city, group = city)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "H", end = 0.8) +
  labs(x = NULL, y = "degC",
       title = "Monthly mean temperature, 2019") +
  theme_calendar()
```

Diurnal profiles by season come from recasting to `s4_h24` and letting
`join_calendar()` attach the timeframe columns (prefixed by the
calendar's name, so several calendars can coexist on one dataset):

```{r diurnal, fig.width = 8, fig.height = 3.2, eval = viz}
panel |>
  recast_calendar(cal, calendars$s4_h24, year = 2019,
                  rule = "weighted_mean") |>
  join_calendar(calendars$s4_h24, timeframes = TRUE) |>
  ggplot(aes(`s4_h24.HOUR`, T10M,
             color = `s4_h24.SEASON`, group = `s4_h24.SEASON`)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~city) +
  scale_color_viridis_d(option = "D", end = 0.85) +
  scale_x_discrete(breaks = calendar_breaks(5)) +
  labs(x = "hour (UTC)", y = "degC", color = "season",
       title = "Seasonal diurnal temperature profiles, 2019") +
  theme_calendar()
```

## Ranges: ribbons

Roll-your-own composition is plain ggplot. Monthly temperature
envelopes — hourly extremes around the mean — are one grouped
`summarise()` and a `geom_ribbon()`:

```{r ribbon, fig.width = 7, fig.height = 3.4, eval = viz}
merra2_cities |>
  mutate(timeslice = datetime_to_timeslice(datetime, calendars$m12)) |>
  summarise(lo = min(T10M), hi = max(T10M), avg = mean(T10M),
            .by = c(city, timeslice)) |>
  mutate(timeslice = factor(timeslice,
                            levels = calendars$m12@members$MONTH)) |>
  ggplot(aes(timeslice, group = city)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = city), alpha = 0.3) +
  geom_line(aes(y = avg, color = city), linewidth = 1) +
  scale_fill_viridis_d(option = "H", end = 0.8) +
  scale_color_viridis_d(option = "H", end = 0.8) +
  labs(x = NULL, y = "degC",
       title = "Temperature envelopes (hourly extremes), 2019") +
  theme_calendar()
```

## Duration curves

The classic energy view: values sorted descending against cumulative
duration. Shares make the x axis honest — `join_calendar(meta = TRUE)`
attaches each timeslice's share of the year, and the cumulative share
times 8760 is hours:

```{r duration, fig.width = 7, fig.height = 3.4, eval = viz}
panel |>
  select(city, timeslice, W50M) |>
  join_calendar(cal, meta = TRUE) |>
  arrange(city, desc(W50M)) |>
  mutate(hours = cumsum(m12_h24.share) * 8760, .by = city) |>
  ggplot(aes(hours, W50M, color = city)) +
  geom_step(linewidth = 0.8) +
  scale_color_viridis_d(option = "H", end = 0.8) +
  labs(x = "hours per year", y = "m/s",
       title = "Wind speed duration curves (m12_h24), 2019") +
  theme_calendar()
```

## Structure figures

The assembled counterparts need no data at all. `calendar_autoplot()`
(also `autoplot()`/`plot()`) draws the calendar's structure as an
icicle, and `calendar_plot()` is the one-call heatmap over the same
layout machinery:

```{r structure, fig.width = 7, fig.height = 3, eval = viz}
calendar_autoplot(calendars$m12_h24)
calendar_plot(calendars$m12_h24, hel_timeslices, palette = "H")
```

Their shared geometry is available without ggplot2 via
`calendar_layout()`; the wall form's frames come from
`calendar_wall_layout()` and `calendar_weekdays()` — all plain data
frames, ready for whatever figure this vignette did not anticipate.

## Fiscal figures

Member order flows into every axis and facet automatically. On the
`fy04_*` calendars (Indian fiscal year: April..March, see
`vignette("calendars")`) the wall facets and every axis start in April
with no plotting changes at all — and because the wall knows the real
dates, its facets carry the Gregorian year (`APR 2019 .. MAR 2020`),
making the rollover explicit. The sample covers calendar 2019, so
fiscal 2019 has its last quarter empty — filter to the fiscal window
first, or Jan-Mar 2019 would land on the Jan-Mar 2020 slots:

```{r fiscal-wall, fig.width = 8, fig.height = 6, eval = viz}
hel_fy <- hel |>
  filter(datetime >= as.POSIXct("2019-04-01", tz = "UTC"))

calendar_wall_plot(calendar("fy04_d365"), hel_fy, z = "SWGDN",
                   year = 2019) +
  scale_fill_viridis_c(option = "H") +
  labs(fill = "W/m2",
       title = "Helsinki solar irradiance, fiscal 2019 (Apr-Mar)")
```

For value axes, label the fiscal months with their calendar year — the
axis runs `Apr 2019 .. Mar 2020`:

```{r fiscal-bars, fig.width = 7, fig.height = 3.4, eval = viz}
fy <- calendar("fy04_m12")

fy_axis <- tibble(timeslice = fy@members$MONTH) |>
  mutate(m = as.integer(sub("^m", "", timeslice)),
         label = paste(month.abb[m], if_else(m >= 4L, 2019L, 2020L)),
         label = factor(label, levels = label))

hel_fy |>
  mutate(timeslice = datetime_to_timeslice(datetime, fy)) |>
  summarise(SWGDN = mean(SWGDN), .by = timeslice) |>
  left_join(fy_axis, by = "timeslice") |>
  ggplot(aes(label, SWGDN)) +
  geom_col(fill = "#1AE4B6") +   # turbo (viridis H) at ~0.35
  scale_x_discrete(drop = FALSE) +   # keep the empty Jan-Mar 2020 slots
  labs(x = NULL, y = "W/m2",
       title = "Helsinki mean solar irradiance, fiscal 2019") +
  theme_calendar(axis.text.x = element_text(angle = 45, hjust = 1))
```

For the spatial dimension, the sibling package geoscales draws region
hierarchies and choropleths the same way (whole-plot
`geoscale_plot()`/`geoscale_autoplot()` over `geoscale_layout()`); see
its "Plotting geoscales" article. `geom_calendar*()` is the stack's
only composable geom family — spatial maps facet and dissolve rather
than layer.
